---
title: "Introduction to asg4nbashots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to asg4nbashots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 4.5,
  out.width = "100%"
)
```

# Overview

The **asg4nbashots** package provides a cleaned NBA Shot Data (1997–2020) and helper functions  
for quick exploration and visualisation, along with a bundled Shiny app for interactive analysis.

The package was inspired by the Stand-up Maths video “[We analysed 4,678,387 NBA shots](https://www.youtube.com/watch?v=yh5c3duQQ1w)”, which discussed how shot selection and expected-points analysis explain modern NBA trends. 

As the video does not publish a dataset or cleaning steps, I use the public “NBA Shot Data 1997–2020” on [data.world/sportsvizsunday](https://data.world/sportsvizsunday/june-2020-nba-shots-1997-2019), covering a similar scope. After cleaning and standardisation, our dataset contains ~4.7M shots (slightly more than 4.68M cited in the video), likely due to differences in source compilation or filtering.

It allows users to:

- Explore millions of NBA shot attempts by distance, zone, and season type  
- Summarise accuracy and expected points trends  
- Launch an interactive app for visual exploration

---

# Load the package and data

``` {r, eval=FALSE}
library(asg4nbashots)

# Load the packaged dataset
data("nba_shots")

# View the structure
dplyr::glimpse(nba_shots)
```

---

# Basic summaries

Here’s how to compute shot accuracy by year and zone using the built-in helper function.

```{r, eval=FALSE}
shot_zone_summary()
```

You can also manually create a quick summary with `dplyr`:

``` {r, eval=FALSE}
library(dplyr)
library(lubridate)

nba_shots |>
  mutate(year = year(game_date)) |>
  group_by(year) |>
  summarise(
    attempts = n(),
    made = sum(shot_made_flag),
    accuracy = made / attempts,
    .groups = "drop"
  )
```

---

# Accuracy and expected points by distance

``` {r, eval=FALSE}
library(ggplot2)

summary_df <- nba_shots |>
  group_by(shot_distance) |>
  summarise(
    accuracy = mean(shot_made_flag),
    expected_points = mean(shot_type * shot_made_flag),
    .groups = "drop"
  )

ggplot(summary_df, aes(shot_distance)) +
  geom_point(aes(y = accuracy), color = "steelblue") +
  geom_point(aes(y = expected_points), color = "red") +
  labs(
    title = "Accuracy and Expected Points by Distance",
    x = "Distance (ft)",
    y = "Average Accuracy / Expected Points"
  ) +
  theme_minimal()
```

---

# Exploring categorical shot zones

You can also look at field-goal accuracy by court zone and season type.

``` {r, eval=FALSE}
nba_shots |>
  group_by(season_type, shot_zone_basic) |>
  summarise(
    attempts = n(),
    made = sum(shot_made_flag),
    accuracy = made / attempts,
    .groups = "drop"
  ) |>
  ggplot(aes(x = shot_zone_basic, y = accuracy, fill = season_type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Shot Accuracy by Zone and Season Type",
    x = "Shot Zone (Basic)",
    y = "Accuracy"
  ) +
  theme_minimal()
[chunk-end]

```

# Launch the interactive Shiny app

``` {r, eval=FALSE}
# run_nbashots_app()
```

The app provides an interactive interface to explore shot location patterns,  
accuracy by player or year, and expected points per distance.

---

# Reproducibility

- Data cleaning pipeline: `data-raw/nba_shots.R`  
- Packaged dataset: `data/nba_shots.rda` (≈18.8 MB)  
- Source: *data.world/sportsvizsunday* (cleaned and extended)

This ensures transparent and reproducible data preparation for analysis and visualisation.

---

# References

- *Tidyverse* packages — data manipulation and plotting  
- *shiny* — interactive app framework  
- *usethis*, *devtools* — package development workflow
