---
title: "Using the Shiny App"
subtitle: "Walkthrough for the NBA Shot Data Explorer (1997–2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Shiny App}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>",
fig.width = 8,
fig.height = 5.2
)
```

# Overview

This vignette shows how to launch and use the interactive Shiny app bundled with **asg4nbashots** to explore shot locations, accuracy, and expected points across the 1997–2020 NBA seasons.

You’ll learn how to:

1. Launch the app from the package
2. Use the sidebar controls (year, season type, outcomes)
3. Read the two tabs (Shot Location Plot; Accuracy & Expected Points)
4. Save/share reproducible settings

---

# 1) Install and load

```{r, eval = FALSE}

# Install (from GitHub)

remotes::install_github("ETC5523-2025/assignment-4-packages-and-shiny-apps-mozuqi25")

# Load the package

library(asg4nbashots)

# Optionally preview the dataset structure

str(nba_shots)
```

---

# 2) Launch the app

Use the exported helper to open the bundled app.

```{r, eval = FALSE}

# Launch the app (opens in a browser window)

asg4nbashots::run_nbashots_app()
```

> Tip: First launch can take a moment while R loads data and compiles packages.

---

# 3) Sidebar controls

The left panel controls the data shown in both tabs.

* **Show All Year Data (Overlay)**
  Displays per-year points on the “Accuracy & Expected Points” tab (you’ll see a cloud of faint dots per distance).

* **Average of Overall Years**
  Shows the pooled average curve across all years (single dot per distance).

* **Year slider**
  When neither of the two checkboxes above is active, the slider filters the app to a specific year.

* **Season type**
  Choose Regular Season, Playoffs, or both.

* **Shot Outcome** (Shot Location tab only)
  Filter to Scored, Missed, or both.
  (On the Accuracy tab, both are enforced since accuracy/expected points require both attempts and makes.)

---

# 4) Shot Location Plot (tab)

What you’ll see:

* Scatter of shots transformed to a single offensive half (left-hand side).
* **Red** = scored, **blue** = missed. Darker areas = more attempts.
* Reference lines mark baseline, half-court and sidelines.

How to use:

1. Toggle **Season type** to compare Playoffs vs Regular Season patterns.
2. Use **Shot Outcome** to focus on scored-only patterns or missed-only clusters.
3. Switch **Year** (or aggregate all years) to see historical changes.

---

# 5) Accuracy & Expected Points (tab)

What you’ll see:

* **Black** points = average shot accuracy by distance (made / attempts).
* **Red** points = average expected points by distance (`mean(shot_type * shot_made_flag)`), so 3s count triple when made.
* The x-axis caps at ~33 ft (≈ 400 in, half court). Grey vertical markers show key distances (e.g., ~2.5 ft, ~22–29 ft).

Typical stories to look for:

* Very close range (< 2.5 ft) has the highest accuracy.
* Mid-range (≈ 2.5–22 ft) maintains fairly steady accuracy.
* Long range (> 22 ft) accuracy drops, but expected points can **rise** near the 3-point line because made shots are worth 3.

Good combos to try:

* **Average of Overall Years** ON, **Show All Year Data (Overlay)** OFF → clean, pooled curves.
* Turn **Overlay** ON → see per-year variability around the pooled curve.
* Turn both OFF and pick a **Year** → one specific season.

---

# 6) Troubleshooting

* **App won’t launch?** Restart R and try `asg4nbashots::run_nbashots_app()` again.
* **Long first load?** That’s normal; the dataset is large. Later loads are faster.
* **Nothing appears / red error in a plot?** Make sure at least one **Season type** is selected.
* **Confusing results?** Check whether you’ve enabled overlays versus pooled averages, or chosen a specific year.

---

# 7) Session info

```{r, eval = FALSE}
sessionInfo()
```

---

# Acknowledgements

This app and dataset packaging were inspired by the Stand-up Maths video *“[We analysed 4,678,387 NBA shots](https://www.youtube.com/watch?v=yh5c3duQQ1w)”*. Source data come from public NBA shot logs (as hosted on [data.world/sportsvizsunday](https://data.world/sportsvizsunday/june-2020-nba-shots-1997-2019)). 

